### 2019-04-18 reviews/icfp-before-rebuttal.txt

ICFP 2019 Paper #111 Reviews and Comments
===========================================================================
Paper #111 TopHat: A formal foundation for task-oriented programming

Review #111A
===========================================================================

Overall merit
-------------
B. OK paper, but I will not champion it.

Reviewer expertise
------------------
Y. I am knowledgeable in the area, though not an expert.

Paper summary
-------------
Interactive workflow "tasks" are a common use of computing
resources. This paper provides a middleweight calculus modeling key
ideas of such interactions: a set of combinators specifies the inputs,
computations, steps, and orderings in such tasks.

The model is a subset of features of existing implementations iTasks
and mTasks, though there's no formal relationship.

The semantics are proven type safe and a small model is implemented in
Idris (though no detail is given on this mechanized model).

The paper concludes with a substantial comparison to related work.

Comments for author
-------------------
The work here is a nice semantics on an important topic that has
received too little attention in the PL community. My score is a B for
several reasons:

- Nothing is _done_ with the semantics beyond sanity checking. Such
  type safety proofs show only that you've built a sensible
  system. What is the semantics for?

- The papers aims and claims don't match up with its actual
  contributions.

  While "formal program verification is important for mission-critical
  software", no actual formal verification happens here---and there's
  no evidence that the workflows described here are mission critical.

  The "_____-oriented programming" framing leads one to compare the
  system to other orientations, which may be unfavorable either
  because other orientations are total views of what programming is
  (and so more substantial ideas than what any single semantics might
  offer) or because other orientations are seen as software
  engineering fads.

- The semantics is entirely operational, which offers the weakest
  reasoning principles as compared with denotational or axiomatic
  models. The semantics and combinators also have gaps, allowing one
  to express senseless workflows.

Despite these flaws, I think the combinators given here will be useful
elsewhere, so I lean towards acceptance.

# What might improve the paper?

- Moderate the claims. The key novelty here is an interesting library
  of combinators in an important domain and a nice semantics for a
  non-recursive fragment. Your properties don't quite "validate" your
  semantics (line 838), but rather serve as a sanity check.

- Use the semantics for something. An Iris-like logic would more than
  suffice for proofs of equivalence of programs. Why not prove that
  one can always eventually eat breakfast? (Such a property is highly
  desirable and sure to be popular.)

  In the related work, you frequently remark that other systems don't
  support formal reasoning---such statements would be more sensible if
  you used your semantics for formal reasoning.

- Summarize the interaction trees of tasks. Throughout the paper, it's
  stressed that TopHat merely specifies collaboration rather than
  explicit communication, as in, e.g., process calculi or session
  types. While such a system is perhaps nice to program in, having a
  term `e : Task t` doesn't tell me much of anything about the set of
  input events `e` can receive. Regular expressions (with embedded
  functional programs to enable transitions) should suffice here, and
  will make for a nice, abstract denotational semantics of tasks.

- Prove that your combinators are orthogonal. Pierce's spider calculus
  (2010) is a nice example: each primitive is proved 'independent' of
  the others.

- Clarify which kinds of tasks are realizable. There's no monadic
  "join" operation that allows us to work with `Task (Task t)`, yet
  the type is expressible. The set of editors is also surely limited:
  what can one do with a $\square (\lambda x:\textsc{Int}. x)$? What
  about $\blacksquare (\textsc{Int}\rightarrow\textsc{Int})$? What
  about $\square (\square 5)$? What about $\blacksquare (\textsc{Task}
  ~ \textsc{Int})$?

  Relatedly, tasks are values in the host language (line 591). What
  eliminators on tasks do you offer?

- You are able to write the smoker program simply because your
  semantics eats all the complexity. What would dining philosophers
  look like in your system? It seems you cannot write a version of it
  that dynamically determines the number of philosophers. What might
  you need to add to the language to support that?

  Later on (line 472) you talk about "immediately reflected" changes,
  but consistent, available (i.e., immediate), and persistent
  propagation is impossible in a distributed system. How should we
  think about the possibility of network delay or failure?
  Communication-based systems force programmers to think about these
  issues, for better or worse.

  Similarly, your comparison to HipHop misses a crucial fact: HipHop's
  Esterel-like semantics has no data races, while TopHat is full of
  data races. While not themselves a problem, there's strong anecdotal
  evidence that racy models are harder to program in.

- Clarify your implementations. What does the Idris implementation do?
  Is it actually executable, or is it a relational model? How does
  your Idris implementation compare to your embedding in iTasks? Have
  you tried testing whether the two semantics behave the same?

- Clarify your relation to FRP. You say that TopHat has no notion of
  time, but $\blacksquare \mathsf{time}$ seems to exactly be a
  time-varying value. It seems that one could use $\blacksquare$ and
  $\RHD$ to write FRP-ish programs, e.g., your derived `delay`
  combinator.

# Questions

You say you "show that the normalisation semantics is a big-step
semantics". Which property is that, and why is it important? Why does
having state complicate a big-step semantics?

Could you encode BPEL in TopHat? If not, which features does BPEL
support that you do not? (E.g., cycles?)

# Minor stuff

- Line 235: What is an invalid event?

- Line 269: What about explicit option types?

- Line 349: We haven't seen these internal operators yet. It might be
  good to at least name them all up front. Could you extend the
  airline example to include them?

- Line 403: You call $\bowtie$ "combination" but perhaps 'parallel
  composition' or 'interleaving' would be more helpful.

- Line 657: It may be worth clarifying that $\rhd$ doesn't need to
  look at $e_2$ because a `C` event is needed. Why doesn't
  $\mathcal{F}$ look at $e_2$ for $\RHD$, though, which needs no
  continue event? Isn't $\blacksquare 5 \RHD \Lightning$ failing?

- Line 793: The typing side conditions on operational are a strange
  violation of the phase distinction.

- Line 949: If iTasks "implicitly has a formal semantics", are there
  any languages that _don't_?

- A few typos: focussing, 'mad an implementation', moddeling

Review #111B
===========================================================================

Overall merit
-------------
B. OK paper, but I will not champion it.

Reviewer expertise
------------------
Z. I am not an expert. My evaluation is that of an informed outsider.

Paper summary
-------------
The paper presents a formalization of the essence of an existing "real-world task" workflow modelling language and methodology called TOP

Comments for author
-------------------
I am an outsider for this particular area of programming languages and was not familiar with TOP.

The paper is well written and interesting, especially for those who know and use TOP and need a detailed description of its formal semantics and some of its properties.

If I have misgivings, I think it is about TOP itself - the question is not its soundness, or its semantics, but its utility for the business domain and the "logical encoding" approach used to formalizing real-world business systems.  My own experience has been with systems such as Oracle Policy Automation which approaches modelling of real-world processes in a vastly different way.  So the sea of symbols as an "encoding" of the real world is imply daunting and unfamiliar to me, and I am skeptical of its utility.  However, since the focus of the paper is on formalization and not on TOP itself, this is me just admitting a (perhaps unfair) bias.

I've given my review a low confidence rating as a result.

Review #111C
===========================================================================

Overall merit
-------------
C. Weak paper, though I will not fight strongly against it.

Reviewer expertise
------------------
Y. I am knowledgeable in the area, though not an expert.

Paper summary
-------------
The paper presents TopHat, an imperative higher-order language with facilities
for interacting with an external environment.

In typical use cases, the environment can be thought of as a user interacting
with the program through a graphical interface. TopHat allows a program to
communicate with the environment via "tasks", which are logical descriptions of
input/output interfaces. Compound tasks can be formed via dedicated combinators:
pairing, monadic-bind-like sequential composition, choice, etc.

Technically, the paper contains an introduction to "task-oriented" programming
in TopHat, a formal semantics for TopHat, a type safety result, and a discussion
of related work.

Comments for author
-------------------

I welcome every effort to make UI programming easier, given the sorry state of
current programming practice. As such, I am largely sympathetic towards the aims
of "task-oriented programming", and find this paper intriguing. Unfortunately,
I'm not convinced that it would be a good fit for ICFP in its current form.

First, I don't think the paper contains enough technical contributions. I
understand that its goal is to give a standalone formalization of task-oriented
programming, which has been around for some time but always in the form of
libraries. But as is, this formalization does not pay off: all applications are
left for future work, which I find highly unsatisfactory. Apparently, we did not
learn anything new about the equivalence of task-oriented programs, nor gain
insight into their implementation, etc. Such points are left for future work,
but would greatly improve the relevance of the work.

Second, I am doubtful of certain design aspects of TopHat.

For example, the language mixes a form of shared memory with bespoke concurrency
constructs, which has several important consequences. In particular, sequential
composition performs speculative execution: if t is a task with a set value, the
sequential composition combinator t ▷ e tries to execute e with this value, and
backtracks if e fails ("failing" is an explicit concept in TopHat),
reestablishing memory as if e had never been executed. This makes composition
very expressive (see e.g. the example given in §3.6, p. 7), but also very
costly, as, in the worst case, the entire store must be duplicated. But I am
even more worried about the impact of this semantics on reasoning: can one
really understand what's going on in a TopHat program which is not very small?
Doesn't this make it hard to understand the complexity of TopHat programs?

Also, while TopHat seems to be a language dedicated to reactive programming,
i.e., continuous interaction with an environment, it does not help with managing
this interaction in a modular way. As far as I see, there is no way for a task A
to interpose itself between the environment and another task B, i.e., for A to
encapsulate B in a way that is transparent from B's perspective. This is
possible in the reactive languages I know of (e.g., Esterel).

Third, in my opinion the paper is poorly written and hard to understand. It
contains too many forward references and makes use of symbols or notions before
they are defined (see below). I also found several statements hard to interpret,
e.g. the task pairing combinator supposedly "introduces interleaving
concurrency" (p. 7), but as far as I see the semantics of TopHat is perfectly
deterministic (given a fixed input sequence) and does not involve interleaved
access to the store. My confusion was exacerbated by the related work section,
where TopHat is compared in vague terms to works whose relevance is not entirely
obvious to me (e.g., process algebra) but where closely-related work from the
ICFP community, such as Carlsson and Hallgren's Fudgets or Conal Elliott's
Tangible Values, is not mentioned at all. Similarly, the comparison to reactive
programming works are not technical enough in my opinion.

## Detailed comments

p. 4

- The "lightning strike" symbol used in the examples should be introduced in the
  text as soon as it is used.

- The "bookFlight ⋈ bookFlight ⋈ bookFlight" example is very problematic. At
  this point, I had zero understanding of TopHat's concurrency model, and thus
  this program looked like it had a data race, making it plainly incorrect. As a
  consequence, I found the paragraphs extolling the virtues of shared-memory
  programming flabbergasting! This should be addressed immediately in the text,
  explaining what is atomic and what isn't.

p. 5

- While discussing the semantics of parallel composition (⋈), it is stated that
  "what is meant by parallel is that the order in which we do tasks and the
  smaller tasks that they are composed of, does not matter". But I cannot
  reconcile this statement with the formal semantics: ⋈ is not commutative at
  all, since for instance "l := True ⋈ l := False" will always result in a store
  containing l ↦ True, and "l := False ⋈ l := True" in one containing l ↦ False.

- "Inputs are synchronous."

  I believe that only a single task may evolve per time step, judging from the
  grammar of inputs presented in Fig 14. So what does this mean, exaclty? That
  program and environment alternate?

p. 6

- I would have liked an intuitive description of the graphical representation of
  choice combinators.

p. 7

- "The details of the solution [Downey's] are not important here."

  Then perhaps one could simplify the preceding paragraph?

p. 8

- Internal choice is mentioned here, without ever being introduced before!

p. 9

- The grammar allows function-valued editors. What does ⊠(Int → Int) look like?

- The grammar allows higher-order tasks and editors. What does ⊠⊠Int look like?

p. 10

- "If e is a reference of type Ref τ, then ▪e is of type Task τ"

  Here τ should be changed into β, or T-Update should be modified.

- The rationale for the external vs. internal distinction is never given. I
  would expect that in an expressive language, external combinators would be
  definable in terms of their internal counterparts.

p. 11

- I liked example 4.4.

p. 13

- "When both branches have a value, it takes the value of the left branch"

  Why not give some combination function?

p. 14

- In Figure 10, F is a function only if ⇓ is deterministic. I believe it is, but
  perhaps this should be stated explicitly somewhere.

p. 15

- Isn't rule S-Xor lacking some premises?

p. 17

- l. 832: obfuscating the link to the public GitHub repository is worse than
  useless. Why not attach an anonymized copy of the source code to the paper?

p. 18

- "we mad": we made.

- "Also, we mad[e] an implementation of TopHat combinators into iTasks. This
  shows that TopHat is a subset of iTasks."

  I'm nitpicking, but I don't like that turn of phrase. Without additional
  information about the simplicity of this implementation, this conclusion is
  incorrect. Otherwise, the fact that you can implement, say, Clean on top of
  x86 assembly would imply that Clean is a subset of x86 assembly.

- l. 926: "understand" -> understanding.

- "We do not know yet if the more complex combinators of iTasks are expressible
  in the basic combinators of TopHat."

  I would have liked to know what these combinators look like.

p. 20

- "However, on IOT devices it is useful to continue running tasks endlessly,
  which is done in mTasks using a "forever" combinator. This is currently not
  possible in TopHat."

  So in TopHat tasks do not "run endlessly", yet they "never terminate" and
  "always keep reacting to events" (p. 6)? This is very confusing.

p. 21

- As I said above, I'm not sure TopHat, a deterministic functional language,
  compares to process algebra.

- l. 996-999: this paragraph is extremely vague ("model collaboration patterns",
  "declarative specification"). Moreover, I don't think comparing "goals" of
  formalisms leads to useful distinctions, otherwise we would conclude that
  functional languages (whose goals are to program computers) and Alonzo
  Church's λ-calculus (whose original goal was to give a foundation for
  mathematics) have nothing to do with each other.

- l. 1015: "syntonisation" -> synchronisation.

- l. 1025: "programatically" -> "programmatically".

- In my opinion, comparing TopHat to ReactiveML would be more relevant, since a
  lot of the HipHop work is about coping with the vagaries of the web
  programming environment.

p. 22

- "FRP and TOP are different systems and have different goals in mind."

  Same remarks as above.

- "TOP has no notion of time" Really? If so, why talk about synchrony and the
  instantaneous character of reactions earlier?

p. 23

- l. 1110: "moddeling" -> "modeling".

Review #111D
===========================================================================

Overall merit
-------------
C. Weak paper, though I will not fight strongly against it.

Reviewer expertise
------------------
Y. I am knowledgeable in the area, though not an expert.

Paper summary
-------------
The paper a formal operational model for a core task-oriented language
inspired by iTasks. The model has a stratified structure. At the bottom
layer there is a base functional typed language with first-class
cells. On top of that there is a language of typed task-related effects.
The semantics of the model is also stratified. At the bottom layer there
is a small-step operational semantics for the base language and on top of
that there are two layers for the effect layer. The bottom of these two
layers deals with task reductions that do not require user input. The
top deals with task reductions that are triggered by user input.
The model is proven type sound and has been implemented in Idris and the
task effect language on top of iTasks.

Comments for author
-------------------
+++++ Comments +++++

++ I think the goal of formalizing the core of iTasks is great. TOP is an
interesting programming paradigm and it would be good to be able to
understand its essence and furthermore to have techniques to prove the
properties of TOP programs.

++ The development of the model seems sensible and the provided examples
are quite interesting.

-- However I think this current paper is far from this goal. As is it
feels more like an exercise in formalisms rather than a research result.
In particular, the approach to describing TOP is a straightforward
adaptation of ideas from process calculi (see work on languages with
session types).

-- Moreover, the task layer and its type system are straightforward
adaptation of monadic effect descriptions. This is a missed opportunity as
a richer type system could be used to prove interesting properties such as
for instance that certain tasks never fail.

-- Finally, there is nothing surprising or challenging in proving type
soundness for the model.

-- This model doesn't meet the requirements to be called a calculus.
Plotkin's seminal paper on the call-by-name and call-by-value lambda
calculus provides a blueprint for how one constructs a calculus for an
existing language. For a more recent example see Florence et al. 's POPL
paper on a calculus for Esterel with the same motivation as this paper but
for Esterel. This effort builds on equatinal theories for effectual programs developed by Felleisen in the 80s. In general, a calculus should allow sound  and consistent
equational reasoning with respect to the existing semantics  of the language
(in this case the graph rewriting one of Clean) and this is possible even in the
presence of effects at the price of weakening reasoning power.

-- Again related to Esterel and the above, the paper ``An axiomatic semantics for
Esterel'' by Simone Tini presents an alternative approach to formalizing
Esterel with the goal of reasoning about Esterel programs. That paper opts
for bisimulation techniques instead of algebraic manipulations of syntax.
Thus similar to TOPHAT, the paper opts for an LTS approach. Again though,
and unlike TOPHAT, the design of the model is directly connected with certain
properties and techniques.

+++++ Small points/typos +++++

page 13: I is mentioned here but it is defined in page 16.

page 18, first line: "we mad" ---> "we made"

page 18, first line: "combinators into" ---> "combinators on top of"

page 21, section 7.3, last line: "syntonisation" ---> "synchronization"

++++++ QUESTIONS THAT THE AUTHORS ARE EXPECTED TO ANSWER ++++++

1) What is challenging/surprising about constructing TOPHAT?

2) Could you please explain what kind of properties of iTasks
programs you intend to intend to try to prove with TOPHAT and
how its design helps this process?

### 2019-04-18 reviews/icfp-rebuttal.md

<!--
Ik probeer zo weinig mogelijk "our" te gebruiken om het wij <-> zij verschil kleiner te maken.
In plaats van "our" gebruik ik "the".

Verder ga ik er bij verschillende dingen wellicht te hard in op de details.
Kunst is om zoveel als mogelijk terug te grijpen naar de dingen die de reviewers positief vinden aan ons paper.
Dat is belangrijk voor jullie om te checken!

Daarbij zit ik nu 26 woorden over de limiet in het algemene deel (dus tot aan het kopje "specific questions").
Iets om rekening mee te houden!

Tekst [tussen blokhaken] ben ik niet tevreden over en zou ik graag anders hebben.

-- TS
-->

# Rebuttal paper #111

We thank the reviewers for their efforts and constructive comments.
We are pleased to read that reviewers consider the paper is "intriguing" and "well written and interesting".
We are glad to hear "the model seems sensible and the provided examples are quite interesting",
and that we are addressing "an important topic that has received too little attention in the PL community".
Also, we will make sure to incorporate all detailed comments in the final version of the paper.

First and foremost, we'd like to apologise naming our model a calculus.
In light of Plotkin's paper from 1973, our model does not meet the requirements to be called a calculus.
Therefore, we will rename our model to a language.
We are grateful for reviewer D to bring this up.

Next to this, reviewers raised three main points:

* The semantics presented is not challenging or special in any way.
* The work does not contain an application of the given semantics
  or a description why it will be usable for formal reasoning.
* Some undesired tasks can be constructed or unwanted effects could happen.

Below we would like to address these three concerns first, after which we answer specific questions of each reviewer.

## Challenge

The challenge in creating this model is indeed not the used techniques,
which are well established and proved their usefulness in computer science.
[The challenge is going back to the roots of workflow and business process modelling,
while adding facilities to specify higher order workflows and generate applications.]
The resulting applications should have automatically generated user interfaces
and no data races despite a multi user environment.

Although the iTasks framework already promises this,
it is a large library [with lots of bells and whistles].
So, in particular, the challenge is to reformulate the iTasks framework and its ideas in a handful of elementary and well-behaved combinators.
The iTasks framework is totally build on top of two "super combinators" called `step` and `parallel`.
Both were created to express every possible collaboration pattern the creators encountered during case studies.
All other simpler combinators are derivations of these two.

For example, `parallel` has the type (in Haskell syntax):
```
parallel
  :: ITask a
  => [ ( ParallelTaskType, ParallelTask a ) ]
  -> [ TaskCont [ ( Int, TaskValue a ) ] ( ParallelTaskType, ParallelTask a ) ]
  -> Task [ ( Int, TaskValue a ) ]
```
Where `ITask` is a type class facilitating (de)serialisability, datatype generic representations,
type information at runtime, and a way to produce user facing editors.
Trimming this combinator down to two elementary combinators (⋈ and ⧫) which,
together with shared memory (■), can model most of its use cases,
is the main challenge solved in this work.

## Application

The analysis that the current work does not apply the semantics is a valid one.
The exclusion of any [proofs] is an intentional one.
In this work we focus on properly introducing the core concepts of TOP,
formally describe them,
and placing this model into the right context by comparing it a large set of other frameworks and algebras.

In an upcoming article, we are using our system to symbolically execute tasks:
"Proving task properties using symbolic TopHat", which is under consideration for IFL'19.
Symbolic execution allows us to prove properties like "breakfast will always be served" (ex 3.1),
and "passengers always get a seat" (ex 2.1).
These kind of properties would not be expressible and provable using a library like the iTasks framework
and need a well described core set of functionality as presented in this work.
Also, symbolic execution allows us to relate traces of inputs to tasks and observations on them,
a suggestion made by reviewer A.

Also, this gives us space to discuss and analyse a bigger case studies for the Dutch tax office and the Dutch coast guard,
which is more mission critical than the examples given in the current work.

## Undesired tasks and effects

### Higher order tasks

As reviewers A and C point out, it is possible to express tasks of higher order types,
which we did not discuss in the paper.
This is an omission which we should fix for the final version of the paper.
Examples given are:

* ⊠(Int -> Int) or □(λx:τ. x)
* ⊠(Task Int) or ⊠(⊠Int)

It is indeed possible to have tasks of functions and tasks of tasks.
On the user interaction part, one could render such editors as a selection menu of functions or tasks already available in the system.
Users could pick one of them, and continue with execution,
such as picking a function to operate on an integer and applying it to a entered integer like such:

t1 := ⊠(Int -> Int) ⋈ ⊠Int ▶︎ λ⟨f,x⟩. □(f x)

Something similar holds for executing a task entered by a user or, more generally, saved by another task.
Here t1 is selected and will be executed in the next normalisation round:

t2 := □(t1) ▶︎ λt. t

This also gives an answer to the missing join operation in the language and is the only way programmers can eliminate tasks.
A top level task can only be eliminated by running the semantics,
as it needs user input.

Saving something else than an basic type in a reference is, however, not allowed by the grammar.
This is a deliberate choice to exclude recursion by using references.
This means the example below are not possible:

* ■l where l : Int -> Int
* ■l where l : Task Int

### Race conditions and distributed systems

Line 363, which states that our model model supports distributed applications,
should never have reached the submitted version.
It is simply not true.
The system can only "generate _multi-user applications_ to support [...] workflows".
Multiple users can interact with an application running on a single server.

There is an important consequence of this choice that we should state more clearly in the paper.
If one takes as a definition for a _data race_ that
_the same program with the same inputs sometimes gives different results_
then this is not possible in TopHat.
The order of execution is completely determined by the order of the inputs,
which we assume to be a serial stream.
Therefore, it is indeed the case that program and environment alternate:
a user gives an input, the system makes some calculations and the user interface will change.
Only then the user can give her next input, etc.

I.e., in the booking example (ex 2.1):
the user who books the seat first gets the seat.
This is exactly the desired behaviour.

## Specific questions

### Speculative evaluation

The speculative evaluation of task continuations in the semantics is needed to express guards.
We acknowledge that taking state into account needs to reestablish memory if the evaluation leads to failure.
A different approach would be move state from the expression layer to the task layer.
We choose not to do this to keep task modelling features and default language features as branching and references completely separate.
We should have mentioned this reasoning in section 5.3.

### Dining philosophers

The dining philosophers are very easy to express using TopHat.
<!-- Definition should be here.  -- TS -->
To express a dynamic version of this problem,
we would need a (restricted) version of recursion.
One could think of built in recursion principle on natural numbers
or a replicate function on tasks.

### 2019-04-18 reviews/icfp-rebuttal.md

<!--
Ik probeer zo weinig mogelijk "our" te gebruiken om het wij <-> zij verschil kleiner te maken.
In plaats van "our" gebruik ik "the".

Verder ga ik er bij verschillende dingen wellicht te hard in op de details.
Kunst is om zoveel als mogelijk terug te grijpen naar de dingen die de reviewers positief vinden aan ons paper.
Dat is belangrijk voor jullie om te checken!

Daarbij zit ik nu zo'n 100 woorden over de limiet van 1000 in het algemene deel (dus tot aan het kopje "specific questions").
Iets om rekening mee te houden!

Ik vraag me even af op welk punt we nog het beste de "de positieve opmerkingen van de reviewers [kunnen] meenemen",
zoals Rinus aangaf:

> TOP is, zoals de reviewers aangeven, een belangrijk concept die een oplossing biedt voor interactieve applicaties (Achilles hiel van FP),
> met iTasks worden complexe commerciele applicaties gemaakt (dus het is praktisch bruikbaar),
> een semantiek die het redeneren over programma's mogelijk maakt is belangrijk, maar niet eenvoudig door de vele mogelijkheden,
> en daarom is TopHat waarbij jullie je bewust beperken een eerste noodzakelijke stap.

Verder zijn er gaten voor een implementatie van de dining philosophers.
Kun jij die toevoegen Markus?
En heb ik de laatste vraag van reviewer D nog niet beantwoord.

@@

# Rebuttal paper #111

We thank the reviewers for their efforts and constructive comments.
We are pleased to read that reviewers consider the paper is "intriguing" and "well written and interesting".
We are glad to hear "the model seems sensible and the provided examples are quite interesting",
and that we are addressing "an important topic that has received too little attention in the PL community".

First and foremost, we'd like to apologise naming our model a calculus.
In light of Plotkin's paper from 1973, our model does not meet the requirements to be called a calculus.
Therefore, we will rename our model to a language.
We are grateful for reviewer D to bring this up.

Next to this, reviewers raised three main concerns:

* The semantics presented is not challenging or special in any way.
* The work does not contain an application of the given semantics.
* Some undesired tasks can be constructed or unwanted effects could happen.

Below we would like to address these three concerns first, after which we answer specific questions of each reviewer.
Naturally, will make sure to incorporate all detailed comments in the final version of the paper.

## General concerns

### Challenge

As reviewers point out,
the challenge in creating this model is indeed not the used techniques.
It is indeed the case that proving type soundness is just a sanity check and not a validation.
This should be reflected in the paper.

<!-- Deze sectie kan iemand vast veeeeel korter herformuleren... ;-)  --TS -->

The main challenge we faced was creating a language which models workflows and business processes,
while being able to specify higher order workflows and generate applications out of such specifications.
The resulting applications should have automatically generated user interfaces
and no data races despite a multi user environment.

Although the iTasks framework already promises this,
it is a large library embedded in an general purpose functional programming language.
So, our challenge was to reformulate the iTasks framework and its ideas in a handful of elementary and well-behaved combinators.

The iTasks framework is totally build on top of two "super combinators" called `step` and `parallel`.
Both were created to express every possible collaboration pattern the creators encountered during case studies.
All other simpler combinators are derivations of these two.

For example, `parallel` has the type (in Haskell syntax):
```
parallel
  :: ITask a
  => [ ( ParallelTaskType, ParallelTask a ) ]
  -> [ TaskCont [ ( Int, TaskValue a ) ] ( ParallelTaskType, ParallelTask a ) ]
  -> Task [ ( Int, TaskValue a ) ]
```
Where `ITask` is a type class facilitating (de)serialisability, datatype generic representations,
type information at runtime, and a way to produce graphical editors.
Trimming this combinator down to two elementary combinators (⋈ and ⧫) which,
together with shared memory (■), can model most of its use cases,
is the main challenge solved in this work.
We did not give this example in section 1.4, where it belongs.

### Application

The analysis that the current work does not apply the semantics is a valid one.
The exclusion of any application is an intentional one.
In this work, we focus on properly introducing the core concepts of TOP,
linking them to a formal description,
showing this description is sane,
and placing the resulting model into the right context by comparing it a fair set of other frameworks and algebras.
These four things lead to a holistic description of TOP which can be used for further research as well as implementation.
Summarising: the presented system is "useful elsewhere", as indicated by reviewer A.

In an upcoming article, we are using our system to symbolically execute tasks.
<!-- Moet zo'n verwijzing er in?  --TS -->
["Proving task properties using symbolic TopHat", under consideration for IFL'19.]
Symbolic execution allows us to prove properties like "breakfast will always be served" (ex 3.1),
and "passengers always get a seat" (ex 2.1).
It will be possible to relate traces of inputs to tasks and observations on them,
a suggestion made by reviewer A.
These kind of properties would not be expressible and provable using a library like the iTasks framework.
They need a well described core set of functionality as presented in this work.

A split also gives us the opportunity to discuss and analyse bigger case studies from the Dutch tax office and the Dutch coast guard.
They are certainly more mission critical than the small examples given in the current work.

### Undesired tasks and effects

#### Higher order tasks

As reviewers A and C point out, it is possible to express tasks of higher order types,
which we did not discuss in the paper.
This is an omission which we should fix for the final version.
Examples given are:

@@

* ⊠(Int -> Int) or □(λx:τ. x)
* ⊠(Task Int) or ⊠(⊠Int)

It is indeed possible to have tasks of functions and tasks of tasks.
On the user interaction part, one could render above editors as a selection menu of functions or tasks already available in the system.
Users could pick one of them, and continue with execution,
such as picking a function to operate on an integer and applying it to a entered integer like such:

@@

This also gives an answer to the missing join operation in the language and is the only way programmers can eliminate tasks.
A top level task can only be eliminated by running the semantics,
as it needs user input.

Saving something else than an basic type in a reference is, however, not allowed by the grammar.
This is a deliberate choice to exclude recursion by using references.
This means the example below are not possible:

* ■l where l : Int -> Int
* ■l where l : Task Int

#### Race conditions and distributed systems

Line 363, which states that our model supports distributed applications,
should never have reached the submitted version.
It is simply not true.
The system can only "generate _multi-user applications_ to support [...] workflows":
multiple users can interact with an application running on a single server.

There is an important consequence of this choice that we should state more clearly in the paper.
If one takes as a definition for a data race that
_the same program with the same inputs sometimes gives different results_
then this is not possible in TopHat.
The order of execution is completely determined by the order of the inputs,
which we assume to be a serial stream.
Therefore, it is indeed the case that program and environment alternate as mentioned by reviewer C:
a user gives an input, the system makes some calculations and the user interface will change.
Only then the user can give her next input, etc.
I.e., in the booking example (ex 2.1):
the user who books the seat first gets the seat.
This is exactly the desired behaviour.

@@

## Specific questions

### Reviewer A

> You say you "show that the normalisation semantics is a big-step semantics".
> Which property is that, and why is it important?
> Why does having state complicate a big-step semantics?

This is an leftover and should not have made it into this version.
We will remove it.

> Could you encode BPEL in TopHat? If not, which features does BPEL
> support that you do not? (E.g., cycles?)

It should be possible to express most BPEL programs in TopHat
as BPEL supports the same basic compositions as TopHat (sequence and parallel)
and the same structured programming constructs.
Only loop constructs are not supported, as TopHat does not have recursion.

An embedding in the other direction, however, would not be possible.
BPEL lacks the higher order constructs which TopHat has.

> You are able to write the smoker program simply because your semantics eats all the complexity.
> What would dining philosophers look like in your system?
> It seems you cannot write a version of it that dynamically determines the number of philosophers.
> What might you need to add to the language to support that?

### Reviewer C

#### On speculative evaluation

@@

#### On speculative evaluation

The speculative evaluation of task continuations in the semantics is needed to express guards.
We acknowledge that taking state into account needs to reestablish memory if the evaluation leads to failure.
A different approach would be move state from the expression layer to the task layer.
We choose not to do this to keep task modelling features and default language features as branching and references completely separate.
We should have mentioned this reasoning in section 5.3.

#### On the notion of time in comparison to FRP

The main difference between time in TopHat and in FRP is that
behaviours in FRP are explicitly functions from time to values.
In TopHat, tasks have a notion of order, but do not explicitly depend on time.

This doesn't mean, however, that task cannot react on a time value, as is done in example 4.5.
Takeaway is that _a clock in TopHat is just a user which updates a shared time value on a regular base_.
We will explicitly mention this in the paper in example 4.5 and in section 7.4.

### Reviewer D

> 1) What is challenging/surprising about constructing TOPHAT?

We hope to have answered this question in the general section,
in the paragraphs about "Challenge".

> 2) Could you please explain what kind of properties of iTasks programs you intend to prove with TOPHAT
>    and how its design helps this process?

<!-- How does the design help??  --TS -->

### 2019-04-24 sections/relatedwork.tex
push fixmes in margin for last proofread
last and only todo
Fixed chorus related work
convert back to two column layout
Hide related work section: related web services

\subsection{Task-oriented programming implementations}

@@

\subsection{Related web services}

\paragraph{IFTTT}

If This Then That (\IFTTT)~\cite{IFTTT} is a web service that allows users to write and
configure simple conditional statements called applets, that can interact with
web services and IoT devices. These applets can be seen as tiny workflows, that
facilitate collaboration between different services and platforms. The
complexity of the flow is very limited however, and is not geared towards humans
collaborating.

\paragraph{Google forms \& SurveyMonkey}

Google Forms~\cite{googleforms} and SurveyMonkey~\cite{wufoo} (and their various other tools like Wufoo, Precise Polling and Zoomerang) are web based survey administration apps that allows
users to compose forms and gather data from them. They are comparable to \TOP and
iTasks in that they provide an easy way for users to construct forms on the web.
However, Google Forms and Wufoo do not allow users to define what the system
then should do with this information, and can therefore not be regarded as a
workflow or business process modelling systems.

\paragraph{Chorus} % (http://www.chorus-home.org/)

Chorus~\cite{chen2017chorus} is a visual programming environment for online
collaboration apps. The goal of Chorus is to allow users to program their own
application, though which a collective task can be coordinated. Users work on a
preset data-type that drives the behaviour of the user defined applications.
\fixme{compare with tophat}

>>>>

\subsection{TOP implementations}

@@

% \subsection{Related web services}
% 
%
% \paragraph{IFTTT}
%
% If This Then That (\IFTTT)~\cite{IFTTT} is a web service that allows users to write and
% configure simple conditional statements called applets, that can interact with
% web services and IoT devices. These applets can be seen as tiny workflows, that
% facilitate collaboration between different services and platforms. The
% complexity of the flow is very limited however, and is not geared towards humans
% collaborating.
%
% \paragraph{Google forms \& SurveyMonkey}
%
% Google Forms~\cite{googleforms} and SurveyMonkey~\cite{wufoo} (and their various other tools like Wufoo, Precise Polling and Zoomerang) are web based survey administration apps that allows
% users to compose forms and gather data from them. They are comparable to \TOP and
% iTasks in that they provide an easy way for users to construct forms on the web.
% However, Google Forms and Wufoo do not allow users to define what the system
% then should do with this information, and can therefore not be regarded as a
% workflow or business process modelling systems.
%
% \paragraph{Chorus} % (http://www.chorus-home.org/)
%
% Chorus~\cite{chen2017chorus} is a visual programming environment for online
% collaboration apps. The goal of Chorus is to allow users to program their own
% application, though which a collective task can be coordinated. Users work on a
% preset data-type that drives the behaviour of the user defined applications.
% Chorus can not be regarded as a fully fledged workflow system, nor does it aim
% to be. The data-type that drives the behaviour of the app does not change during
% execution, and therefore heavily limits what flows can be encoded.
